Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model

### 2023-03-08 src/docs_bayescomp-poster/bayescomp-poster.Rmd
Work on poster for BayesComp

---
main_topsize: 0.18
main_bottomsize: 0.03
poster_height: "48in"
poster_width: "36in"
font_family: "Open Sans"
# Essentials
title: Integrated nested Laplace approximations for extended latent Gaussian models, with application to the Naomi HIV model
author:
  - name: '**Adam Howes**'
    affil: 1, 2
    main: true
    twitter: adamhowes
    email: ath19@ic.ac.uk
  - name: "Alex Stringer"
    affil: 3
  - name: "Seth R. Flaxman"
    affil: 4
  - name: "Jeffrey W. Eaton"
    affil: 2
affiliation:
  - num: 1
    address: Department of Mathematics, Imperial College London
  - num: 2
    address: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
  - num: 3
    address: Department of Statistics and Actuarial Science, University of Waterloo
  - num: 4
    address: Department of Computer Science, University of Oxford
main_findings: "Simplified INLA, automatic differentiation and adaptive Gauss-Hermite quadrature for fast and accurate approximate inference"
logoleft_name: "IMP_ML_W_N.png"
# logoright_name: "UNAIDS-reverse-EN.gif"
# Style & formatting
column_numbers: 3
title_textsize: "60px"
author_textsize: "1.17em"
authorextra_textsize: "35px"
affiliation_textsize: "25px"
affiliation_textcol: '#00000080'
caption_fontsize: "20px"
# Main
main_fontfamily: "Open Sans"
# Poster body options
primary_colour: '#38761C'
secondary_colour: '#3b77D8'
accent_colour: "#5d19c4"
body_bgcol: "#ffffff"
body_textsize: "35px"
body_textcol: "#000000"
reference_textsize: "25px"
sectitle_textsize: "40px"
# Standard options
bibliography: citations.bib
output: 
  posterdown::posterdown_betterport:
    self_contained: false
    pandoc_args: --mathjax
    highlight: espresso
    number_sections: false
---

<!-- Using LaTeX to create spacing -->
$$\\[0.15in]$$

## The Naomi model

$$\\[0.15in]$$

@@

### Laplace approximation

### Adaptive Gauss-Hermite Quadrature

Approximate integrals by 
$$
\int_\Theta p(\theta) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p(\hat \theta + Lz) \omega(z)
$$
with **Gauss-Hermite quadrature** rule $z \in \mathcal{Q}(m, k)$ **adapted** based upon the mode $\hat \theta = \text{argmax}_{\theta \in \Theta} \in p(\theta)$ and lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$ of the target.

### Our algorithm

Given C++ user template for $- \log p(y, x, \theta)$:

## Comparison

## Conclusions

>>>>

---
poster_height: "46.8in"
poster_width: "33.1in"
font_family: "Open Sans"
# Essentials
title: Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model
# titlebox_height: "10%"
author:
  - name: "Adam Howes"
    affil: 1, 2
    main: true
    email: ath19@ic.ac.uk
  - name: "Alex Stringer"
    affil: 3
  - name: "Seth R. Flaxman"
    affil: 4
  - name: "Jeffrey W. Eaton"
    affil: 2
affiliation:
  - num: 1
    address: Department of Mathematics, Imperial College London
  - num: 2
    address: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
  - num: 3
    address: Department of Statistics and Actuarial Science, University of Waterloo
  - num: 4
    address: Department of Computer Science, University of Oxford
logoleft_name: "mrc-gida-icl-uw-uo.png"
# Style & formatting
column_numbers: 3
title_textsize: "75px"
author_textsize: "50px"
authorextra_textsize: "35px"
affiliation_textsize: "35px"
caption_fontsize: "20px"
# Title box
title_textcol: "#000000"
author_textcol: "#000000"
affiliation_textcol: "#000000"
# Main
main_fontfamily: "Open Sans"
# Poster body options
primary_colour: "#ffffff"
secondary_colour: "#000000"
accent_colour: "#000000"
body_bgcol: "#ffffff"
body_textsize: "35px"
body_textcol: "#000000"
reference_textsize: "25px"
sectitle_textsize: "40px"
# Standard options
bibliography: citations.bib
output: 
  posterdown::posterdown_html:
    self_contained: false
    pandoc_args: --mathjax
    highlight: espresso
    number_sections: false
---

```{css, echo=FALSE}
div.logo_left{
  width: 35%;
}
div.poster_title{
  width: 60%;
}

```

## Summary

* Approximate Bayesian inference method combining Laplace approximations and adaptive Gauss-Hermite quadrature 
* Motivated by an evidence synthesis model for small-area estimation of HIV indicators in sub-Saharan Africa
* Implemented as a part of the `aghq` R package, allowing flexible use for any model with a `TMB` C++ user template

<!-- 38761C -->

## The Naomi HIV model

* District-level model of HIV indicators synthesising data from household surveys, antenatal care (ANC) clincs, and routine service provision of antiretroviral therapy (ART) [@eaton2021naomi]
* Yearly estimation process: model run interactively by country teams using the web-app `naomi.unaids.org`
* Inference conducted in minutes using empirical Bayes (EB) and a Gaussian approximation via Template Model Builder `TMB` [@kristensen2016tmb]
* Would take days to get accurate answers with MCMC via `tmbstan` [@monnahan2018no]
* Motivates looking for a fast, approximate approach

## Extended latent Gaussian models

* Three stage hierarchical model with observations $y$, latent field $x$ and hyperparameters $\theta$
* In a latent Gaussian model (LGM) [@rue2009approximate] the conditional mean depends on exactly one structured additive predictor $\mu_i = g(\eta_i)$ with $g: \mathbb{R} \to \mathbb{R}$
  * The `R-INLA` implementation of integrated nested Laplace approximations applies only to LGMs
* Extended latent Gaussian models (ELGM) remove this requirement such that $\mu_i = g(\eta_{\mathcal{J_i}})$ where $g_i: \mathbb{R}^{|\mathcal{J_i}|} \to \mathbb{R}$
* Naomi is an ELGM, not an LGM, because we model:
  * ANC indicators offset from household survey indicators
  * Incidence as depending on prevalence and ART coverage
  * Observed data as aggregated finer processes
  * Individuals as possibly attending ART clinics outside their home district
* ELGM precision matrices less sparse than those of LGMs

@@

* Laplace approximation: Integrate out variables using a Gaussian approximation to the denominator
$$
p(\theta, y) \approx \tilde p_\text{LA}(\theta, y) = \frac{p(y, x, \theta)}{\tilde p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)}
$$
where $\tilde p_\text{G}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x(\theta), \mathbf{H}(\theta)^{-1})$
* Adaptive Gauss-Hermite Quadrature:
$$
\int_\Theta p(\theta) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p(\hat \theta + Lz) \omega(z)
$$
where the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(m, k)$ is adapted based upon the mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p(\theta)$ and lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$ of the target
* Given a C++ user template for $- \log p(y, x, \theta)$ our algorithm is summarized by Figure 1

```{tikz, tikz-ex, echo=FALSE, fig.align='center', fig.cap = "Flowchart describing the algorithm", fig.ext = 'png', cache=TRUE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=4cm, auto,>=latex', thick, scale = 1]
\node (1) {$p(\theta, x, y)$};
\node (2) [right of=1] {$\tilde p_\text{LA}(\theta, y)$};
\node (3) [below of=1] {$\tilde p_\text{LA}(x_i, \theta, y)$};
\node (4) [below of=2] {$\tilde p_\text{AQ}(y)$};
\draw[->] (1) to node {Laplace} (2);
\draw[->] (1) to node {Laplace} (3);
\draw[->] (2) to node {Quadrature} (4);
\end{tikzpicture}
```

## Application to Malawi data

```{r fig2, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Model outputs for adults 15-49 in January 2016.'}
knitr::include_graphics("fig2.png")
```

* $\text{dim}(x)$ is 491 and $\text{dim}(\theta)$ is 24
* Figure \@ref(fig:fig2) illustrates example model outputs: HIV prevalence, ART coverage, HIV incidence, and number of new infections, at the district level

```{r fig3, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Inference reults and ECDF comparison for one $x_i$.'}
knitr::include_graphics("fig3.png")
```

* Compare inferences from `TMB`, `aghq`, `adam` and `tmbstan` using Kolmogorov-Smirnov (KS) tests on hyperparameter, latent field, and output posterior marginals
* Figure \@ref(fig:fig3) illustrates the ECDF difference between `TMB` and `adam` to `tmbstan` for one particular node in the model

## Future directions

* Scaling up the hyperparameter grid beyond EB
  * Any dense grid would be $k^{24}$ i.e. impractical
  * Sparse grids? Good subspaces of $\mathbb{R}^{24}$?
* Add Laplace matrix algebra approximations [@wood2020simplified]
* More comprehensive inference comparison e.g. maximum mean discrepancy, Pareto-smoothed importance sampling

<font size="6"> **Interested to read more?** Working notebooks and R code available from `github.com/athowes/multi-agyw`. </font>
